---
title: "Forecast daily bike rental demand using time series models"
date: "`r Sys.Date()`"
output: html_document
author: "Hitesh Jangid"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on forecasting daily bike rental demand using time series models in R. It contains analysis such as data exploration, summary statistics and building the time series models. The final report was completed on `r date()`. 

**Data Description:**

This dataset contains the daily count of rental bike transactions between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information.

**Data Source:** https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset

**Relevant Paper:** 

Fanaee-T, Hadi, and Gama, Joao. Event labeling combining ensemble detectors and background knowledge, Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg



# Task One: Load and explore the data

## Load data and install packages

```{r}

## Import required packages
library("tidyverse")
library("ggplot2")
library("dbplyr")
library("timetk")
library("tseries")
library("forecast")

```


## Describe and explore the data

```{r}

# Dataset Source
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00275/Bike-Sharing-Dataset.zip"

# Download the zip file
temp <- tempfile()
download.file(url, temp)

# Unzip the file
unzip(temp, "day.csv")

# Read the csv file
day_data <- read.csv("day.csv")

# Clean up
unlink(temp)

```


```{r}

# Explore data
day_data[,"dteday"] <- as.Date(day_data[,"dteday"])
day_data[,"ncnt"] <- day_data[,"cnt"] / max(day_data[,"cnt"])
day_data[,"nr"] <- day_data[,"registered"] / max(day_data[,"registered"])
day_data[,"rr"] <- day_data[,"cnt"] / max(day_data[,"registered"])
summary(day_data)

```



# Task Two: Create interactive time series plots

```{r}
## Read about the timetk package

day_data %>% group_by(yr) %>% plot_time_series(dteday, temp, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Temperature", 
                              .title = "Normalized Temperature vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, hum, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Humidity", 
                              .title = "Normalized Humidity vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, windspeed, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Windspeed", 
                              .title = "Windspeed vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, ncnt, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Bike Rentals", 
                              .title = "Normalized Bike Rentals vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, nr, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Registered Rentals", 
                              .title = "Normalized Registered Rentals vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, rr, .color_var = season, .x_lab = "Date", .y_lab = "Ratio", 
                              .title = "Ratio of Rentals to Registration vs Date for Day Data", .interactive = TRUE)
```




# Task Three: Smooth time series data

```{r}
#since the humdity and windspeed data has so much noise, I will focus on temperature and bike rentals
#data cleaning
day_data[,"temp"] <- tsclean(day_data[,"temp"])
day_data[,"ncnt"] <- tsclean(day_data[,"ncnt"])
day_data[,"nr"] <- tsclean(day_data[,"nr"])
day_data[,"rr"] <- tsclean(day_data[,"rr"])
head(day_data)
```


```{r}
#day_data
day_data %>% group_by(yr) %>% plot_time_series(dteday, temp, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Temperature", .title = "Normalized Temperature vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, ncnt, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Bike Rentals", .title = "Normalized Bike Rentals vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, nr, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Registered Rentals", .title = "Normalized Registered Rentals vs Date for Day Data", .interactive = TRUE)
```


```{r}
day_data %>% group_by(yr) %>% plot_time_series(dteday, rr, .color_var = season, .x_lab = "Date", .y_lab = "Normalized Registered Rentals", .title = "Ratio of Rentals to Registration vs Date for Day Data", .interactive = TRUE)
```



# Task Four: Decompose and assess the stationarity of time series data

```{r}
#day data
day_data[,"temp"] %>% adf.test()
```


```{r}
day_data[,"ncnt"] %>% adf.test()
```


```{r}
day_data[,"nr"] %>% adf.test()
```


```{r}
day_data[,"rr"] %>% adf.test()
```


```{r}
#decomposes the data
freq <- 365

#normalized day rentals
norm_rentals <- ts(day_data[, "nr"], frequency = freq)
decomped1 <- stl(norm_rentals, "periodic")
plot(decomped1$time.series[,2], ylab = "Stationary of the Normalized Rental Reservations", 
     xlab = "Day of the Year")
```


```{r}
checkresiduals(decomped1$time.series[, 3])
```


```{r}
#normalized day counts
norm_cnt <- ts(day_data[, "ncnt"], frequency = freq)
decomped2 <- stl(norm_cnt, "periodic")
plot(decomped2$time.series[,2], ylab = "Stationary of the Normalized Rental Counts", 
     xlab = "Day of the Year")
```


```{r}
checkresiduals(decomped2$time.series[, 3])
```


```{r}
#normalized day rental rates
norm_rr <- ts(day_data[, "rr"], frequency = freq)
decomped3 <- stl(norm_rr, "periodic")
plot(decomped3$time.series[,2], ylab = "Stationary of the Normalized Rental Counts to Reservations", 
     xlab = "Day of the Year")
```


```{r}
checkresiduals(decomped3$time.series[, 3])
```


```{r}
#returns the stats
print("-------Noramlized Rental Reservations")
```


```{r}
shapiro.test(decomped1$time.series[, 3])
```


```{r}
print("-------Normalized Count of Rentals")
```


```{r}
shapiro.test(decomped2$time.series[, 3])
```


```{r}
print("-------Normalized Ratio of Rentals to Reservations")
```


```{r}
shapiro.test(decomped3$time.series[, 3])
```



# Task Five: Fit and forecast time series data using ARIMA models

```{r}
#bike count predictions
fit1 <- auto.arima(norm_cnt, seasonal = TRUE, )
hist(fit1$residuals, xlab = "Residual", ylab = "Distribution", main = "Histogram of Model Errors - Bike Count")
```


```{r}
shapiro.test(fit1$residuals)
```


```{r}
prediction1 <- forecast(fit1, 25)
plot(prediction1, xlab = "Date", ylab = "Normalized Count of Rentals", main = "Prediction of Bike Rental Counts")
```


```{r}
#predictions of number of reservations
fit2 <- auto.arima(norm_rentals, seasonal = TRUE, )
hist(fit2$residuals, xlab = "Residual", ylab = "Distribution", main = "Histogram of Model Errors - Rental Count")
```


```{r}
shapiro.test(fit2$residuals)
```


```{r}
prediction2 <- forecast(fit2, 25)
plot(prediction2, xlab = "Date", ylab = "Normalized Rentals", main = "Prediction of Bike Rentals")
```


```{r}
#bike count predictions
fit2 <- auto.arima(norm_cnt, seasonal = TRUE, )
hist(fit2$residuals, xlab = "Residual", ylab = "Distribution", main = "Histogram of Model Errors - Count to Rental Ratio")
```


```{r}
shapiro.test(fit2$residuals)
```


```{r}
prediction3 <- forecast(fit2, 25)
plot(prediction3, xlab = "Date", ylab = "Normalized Rental Ratio", main = "Prediction of Bike Rentals to Reservations")
```



# Task Six: Findings and Conclusions
After processing the raw data and using the ARIMA package to model ride share data, I was able to make predictions for the 25 days beyond the current data set. Qualitatively the data shows that when the weather gets warmer the number of bike rentals increase, and over the course of two years the number of rentals increases over the number of rentals from the previous year. As the data terminates at the end of one cycle, I expect the number of rentals to increase to a level higher than it was a year before, which is what the models are predicting. Therefore the results were what I expected - the data appears to oscillate up and down over a 1 year period with the overall data moving towards higher rental numbers.































